Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-30 icfp.tex
finished section 5

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

Datatype declarations $\datatype{T~\overline{p}}{\overline{K:\overline{\sigma
\to_\pi}~T~\overline{p} }}$ involve the name of the type being declared $T$
parametrized over multiplicity variables $\ov{p}$, and a set of the data
constructors $K$ with signatures indicating the type and multiplicity of the
constructor arguments. Note that a linear resource is used many times when a
constructor with an unrestricted field is applied to it, since, dually, pattern
matching on the same constructor with an unrestricted field allows it to be
used unrestrictedly. Programs are a set of declarations and a top-level
expression.

@@

\noindent Finally, unrestricted resources are equivalent to $\D$-bound
variables with an empty ($\cdot$) usage environment.

@@

\TypePreservationTheorem
%
\ProgressTheorem
% \noindent Type preservation states that a well-typed expression $e$ that
% evaluates to $e'$ remains well-typed under the same context:
%
Type preservation
% ($\S$~\ref{sec:proof:type-preservation})
follows by structural induction on the reductions $e \longrightarrow
e'$ from the operational semantics. Most cases are straightforward and usually
appeal to one or more of the substitution lemmas described below. The most
interesting case is that of case expressions whose scrutinee can be further
evaluated -- we branch on whether the scrutinee becomes in WHNF, and invoke the
\emph{Irrelevance} lemma if so.
%
% This case guarantees that the separation of rules for treating scrutinees is
% consistent, in the sense that a well-typed case expression with a scrutinee not
% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
%
%
%
%\noindent
% Progress states that the evaluation of a well-typed term does not block:
% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).

@@

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}

One of the primary goals of the Linear Core type system is being suitable for
intermediate representations of optimising compilers for lazy languages with
linear types. In light of this goal, we prove that \emph{multiple optimising
transformations} are type preserving in Linear Core, and thus preserve linearity.

% The optimising transformations proved sound wrt Linear Core in this section
% have been previously explained and motivated in
% Section~\ref{sec:core-to-core-transformations}.
% %
Transformations are described by an arbitrary well-typed expression with a certain shape, on
the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
the right hand side (rhs) that we prove to be well-typed.
%
% For our proofs, we assume the lhs to be a well-typed expression and prove the
% rhs is well-typed as well.
%
For each transformation, we describe the intuition behind the transformation
preserving linearity in our system.

\todo[inline]{Aqui uma lista itemizada das transformações todas. As
  provas todas para apendice.}

% \subsubsection{Inlining}

@@

\subsubsection{Reverse Binder Swap Considered Harmful\label{sec:reverse-binder-swap-considered-harmful}}

The reverse binder swap transformation substitutes occurrences of the case
binder $z$ in case alternatives by the scrutinee, when the scrutinee is a
variable $x$.

\ReverseBinderSwapTheorem

\noindent This is exactly reverse from what the binder swap transformation
does in hope of eliminating multiple uses of $x$ so as to inline it. However, by
using the scrutinee $x$ instead of the case binder, we might be able to float
out expressions from the alternative using the case binder. For example, we
might float an expensive computation involving $z$ out of the case alternative,
where $z$ is out of scope but $x$ isn't:
\[
\begin{array}{l}
\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~z) \dots\} }}{\dots}\\
\Longrightarrow_\textrm{Reverse binder swap}\\
\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~x) \dots\} }}{\dots}\\
\Longrightarrow_\textrm{Float out}\\
\lambda x.~\llet{t = expensive~x}{\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots t \dots\} }}{\dots}}\\
\end{array}
\]
If $go$ is a loop, by applying the reverse binder swap we now only compute
$expensive~x$ once instead of in every loop iteration.

Despite GHC applying the reverse binder swap transformation to core programs
during Core-to-Core optimisation passes, this optimisation violates linearity
when considered as a transformation on Linear Core programs. In practice, the
optimisation preserves linearity in Core when applied as part of the GHC
transformation pipeline only due to the occurrence analyser being naive with
regard to semantic linearity. Initially, it might seem as though an expression
in which a variable $x$ occurs both in the case scrutinee and in the
alternatives is linear, for example:
\[
\G; \xl \vdash \ccase{x}{\_ \to x} : \s
\]
The reasoning is done by branching on whether $x$ refers to an expression in
WHNF or an unevaluated thunk.
\begin{itemize}
\item If $x$ refers to an unevaluated expression, then scrutinizing it results
in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
\item If $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables)
\end{itemize}
%
% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
%
Even though, on its own, it makes intuitive sense that this example indeed uses
$x$ linearly, when considered as part of a complete type system, allowing this
expression to be linear makes the system unsound.

We recall that $\beta$-reduction reduces an application of a linear function
using call-by-name -- if we know the argument is used exactly once, a binding
to share the result of computing the argument is unnecessary, so we instead
substitute the argument expression for the linearly-bound variable in the
$\lambda$-body directly.

Consider the function application $$(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$$
where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
is a linear function by the reasoning above, $\beta$-reduction transforms the
application in $(\ccase{x}{\_ \to x})[use~y/x]$, i.e. $\ccase{use~y}{\_ \to
use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
expression, yet it occurs in the body of the case alternative as well --
linearity is violated by using the linear resource $y$ twice.

Essentially, Linear Core would be unsound, and even duplicate resources, if the
above kind of expressions, where linear variable scrutinees occur in the
alternatives body, were well-typed, because of its interaction with the
call-by-name reduction of linear functions. In this sense, the reverse binder
swap is an optimisation that creates ill-typed expressions from well-typed
ones, so it is deemed an invalid optimisation that doesn't preserve types in
our system.

The reverse binder swap is not a problem in the GHC simplifier because of the
weaker notion of linearity understood by occurrence analysis. Occurrence
analysis is a static analysis pass which can be used to determine whether a
lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
$\beta$-reduction is done with call-by-need on such an expression. If the above
example were reduced with call-by-need:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBNeed}
\llet{x = use~y}{\ccase{y}{\_ \to y}}
\end{array}
\]
Then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
variable, which is indeed an expression semantically linear in $x$.

Concluding, in being able to understand more programs as linear, our type
system allows more expressions to be considered linear for $\beta$-reduction
without a let-allocation, however, it makes reverse binder swap an invalid
transformation since its output, when considered linear, might violate
linearity when further optimised.

@@

\section{Optimisations preserve linearity}

>>>>

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

Datatype declarations $\datatype{T~\overline{p}}{\overline{K:\overline{\sigma
\to_\pi}~T~\overline{p} }}$ involve the name of the type being declared $T$
parametrized over multiplicity variables $\ov{p}$, and a set of the data
constructors $K$ with signatures indicating the type and multiplicity of the
constructor arguments. Note that a linear resource is used many times when a
constructor with an unrestricted field is applied to it, since, dually, pattern
matching on the same constructor with an unrestricted field allows it to be
used unrestrictedly.
%Programs are a set of declarations and a top-level
%expression.

@@

\noindent Finally, unrestricted and $\D$-bound
variables with an empty usage environment are equivalent.

@@

\TypePreservationTheorem
%
\ProgressTheorem
% \noindent Type preservation states that a well-typed expression $e$ that
% evaluates to $e'$ remains well-typed under the same context:
%
Type preservation
% ($\S$~\ref{sec:proof:type-preservation})
follows by structural induction on the reductions $e \longrightarrow
e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
interesting case is that of case expressions whose scrutinee can be further
evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
%
% This case guarantees that the separation of rules for treating scrutinees is
% consistent, in the sense that a well-typed case expression with a scrutinee not
% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
%
%
%
%\noindent
% Progress states that the evaluation of a well-typed term does not block:
% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).

@@

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
A main goal of the Linear Core type system is 
to serve as a typed intermediate representation for optimising
compilers of lazy languages with linearity. In light of this,
we show that multiple optimising transformations are type
preserving in Linear Core (i.e.,~preserve linearity semantically).

% The optimising transformations proved sound wrt Linear Core in this section
% have been previously explained and motivated in
% Section~\ref{sec:core-to-core-transformations}.
% %
%\{\overline{\rho\to e'}\}
\begin{figure}[t]

  {\small
  \[
    \begin{array}{ll}
   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
                                  \mbox{Case of known constr.}\\
      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
            \Longrightarrow 
      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
      & \mbox{Case-of-case}\\
      (\lambda \y[\pi].~\llet{\xD = e}{e'})
      \Longrightarrow
      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
                               &
                                 \mbox{Commuting let-case}\\
      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1], \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{with $x$ fresh}) \\
     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[t], \overline{\rho_k\to e_k}\}} 
                                                                        & \mbox{Commuting case-let}\\
      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
                                                                      & \mbox{Commuting let-let}\\
      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
                                                                      & \mbox{Binder swap}\\
      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
      %                                                                 & \mbox{Reverse binder swap}
    \end{array}
  \]}
  \caption{Optimising transformations validated by Linear Core\label{fig:opttrans} (see Appendix~\ref{app:optimisations} for details)} 
\end{figure}

We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
where $e_1$ is an arbitrary \emph{well-typed} expression
matching a certain shape (or of a certain type), which is
transformed into expression $e_2$. Validating a transformation
entails showing that $e_2$ is well-typed. We often annotate the arrow
with the name of the transformation when describing chains of such
transformations.

The list of transformations we have
validated is given in Figure~\ref{fig:opttrans}.
%
The behaviour of most transformations can be inferred from its name:
Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
and case of known constructor effectively perform evaluation of the given expression;
case-of-case commutes cases in scrutinee position inside the outer case alternatives;
let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
the four commuting conversions allow let-bound expressions to be commuted with other
Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
finally, binder swap substitutes occurrences of a scrutinee variable for the
case binder. While many of the transformations above are not obviously optimising,
they are chained in the compiler pipeline to expose optimisation opportunities.

% Transformations are described by an arbitrary well-typed expression with a certain shape, on
% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
% the right hand side (rhs) that we prove to be well-typed.
% %
% % For our proofs, we assume the lhs to be a well-typed expression and prove the
% % rhs is well-typed as well.
% %
% For each transformation, we describe the intuition behind the transformation
% preserving linearity in our system.

% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
%   provas todas para apendice.}

%%% COMENTADO AQUI
% \subsubsection{Inlining}

@@

\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}

We now discuss an issue we have found with the so-called reverse binder swap
transformation.
%
%
Reverse binder swap is dual to the binder swap transformation,
substituting occurrences of the case
binder $z$ by the scrutinee in the special case when the scrutinee is a
variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By
using the scrutinee $x$ instead of the case binder, we might be able to float
out expressions from the alternative using the case binder. For example, we
might float an expensive computation involving $z$ out of the case alternative,
where $z$ is now out of scope:

{\small\[
\begin{array}{l}
(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
\Longrightarrow_\textrm{Reverse binder swap}\\
(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
\Longrightarrow_\textrm{Commuting case-let}\\
\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
\end{array}
\]}

\noindent If $go$ is compiled into a loop, we only compute
$\mathit{expensive}~x$ once instead of in every loop iteration.

Despite GHC applying the reverse binder swap transformation 
during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
when considered as a transformation on Linear Core programs. In practice, the
optimisation preserves linearity when applied as part of the GHC
pipeline due to the occurrence analyser being naive with
regard to semantic linearity. Initially, it might seem as though an expression
in which a variable $x$ occurs both in the case scrutinee and in the
alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
  If $x$ refers to an unevaluated expression, then scrutinizing it results
in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
%
On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables).
%\end{itemize}
%
% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
%

Even though, on its own, it makes intuitive sense that this example indeed uses
$x$ linearly, when considered as part of a complete type system, allowing this
expression to be linear makes the system \emph{unsound}.
To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
using call-by-name -- if we know the argument is used exactly once, a binding
to share the result of computing the argument is unnecessary, so we instead
substitute the argument expression for the linearly-bound variable in the
$\lambda$-body directly.

Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
is a linear function, $\beta$-reduction transforms the
application into $\ccase{use~y}{\_ \to
  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
that also occurs in the case alternative --
linearity is violated by using the linear resource $y$ twice.
%
% Essentially, Linear Core would be unsound, and even duplicate resources, if the
% above kind of expressions, where linear variable scrutinees occur in the
% alternatives body, were well-typed, because of its interaction with the
% call-by-name reduction of linear functions.
In this sense, the reverse binder
swap is an optimisation that creates ill-typed expressions from well-typed
ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
our system.

Reverse binder swap is not a problem in the GHC simplifier because of the
weaker notion of linearity understood by occurrence analysis. Occurrence
analysis is a static analysis pass which can be used to determine whether a
lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
$\beta$-reduction is done with call-by-need on such an expression. If the above
example were reduced with call-by-need
% \[
% \begin{array}{l}
% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
% \Longrightarrow_\textrm{CBNeed}
% \llet{x = use~y}{\ccase{y}{\_ \to y}}
% \end{array}
% \]
then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
variable, which is indeed semantically linear in $x$.

% Concluding, in being able to understand more programs as linear, our type
% system allows more expressions to be considered linear for $\beta$-reduction
% without a let-allocation, however, it makes reverse binder swap an invalid
% transformation since its output, when considered linear, might violate
% linearity when further optimised.

@@

\section{Optimisations preserve linearity}\label{app:optimisations}
